Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes

### 2020-01-14 ms/manuscript.Rmd
Update MS
- Fix spacing when gobbled after xspace
- Use single quotes throughout (New Phyt style)
- Don't use equations
- Only use siunitx functions supported by pandoc

@@

```{r word-count table}
tibble(
  words = c(
    "Total word count (Introduction, Materials and Methods, Results, Discussion):_n",
    "Summary:_n",
    "Introduction:_n",
    "Materials and Methods:_n",
    "Results:_n",
    "Discussion:_n",
    "Acknowledgements:_n"),
  figs_tabs = c(
    "No. of figures:_7 (all in color)",
    "No. of tables:_5",
    "No. of supporting information files:_8 (Figs. S1-S2; Tables S1-S6)", "", "", "", "")
) %>%
  separate(words, c("words", "count"), sep = "_") %>%
  separate(figs_tabs, c("figs_tabs", "n"), sep = "_", fill = "right") %>%
  mutate(n = replace_na("")) %>%
  gt() %>%
  cols_label(
    words = "",
    count = "",
    figs_tabs = "",
    n = "",
  ) %>%
  as_rtf() %>%
  write_lines(here::here("ms/word_count.rtf"))
```

@@

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic fundamentally changed climates and had an immense effect on terrestrial ecosystems, including increased global precipitation and the creation of novel niches [@Niklas1983; @Kenrick1997a; @Lloyd2008; @Boyce2010]. 
The heterogeneous niche space of the angiosperm canopy comprising stratified light and humidity levels and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. 
The transition to epiphytic growth was particularly important for modern fern diversification: \circa 29% of all fern species are epiphytes vs. \circa 10% across all vascular plants [@Kress1986].
Moreover, major transitions to epiphytism occurred at least five times within ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012].
This has led to the hypothesis that epiphytism was a key factor contributing to extant fern diversification [@Schneider2004d; @Schuettpelz2009; @WatkinsJr.2012].
Recent studies exploring this idea [@Sundue2015; @Testo2016a; @Lehtonen2017], however, have largely failed to identify evidence of rapid diversification associated with epiphytism in ferns. 
This raises the possibility that modern fern diversification was not shaped by the evolution of epiphytism *per se*, but rather by the exploitation of various niches within the canopy and on the forest floor.

@@

To gain insight into the role of the canopy in promoting or restricting diversity, a functional and community-based approach can be applied to study this question. 
If the canopy is uniformly harsher than the terrestrial environment, thus acting as a "filter", we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to increased functional diversity of epiphytic communities. 
Furthermore, the role of the canopy as a promoter or restrictor of diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kromer2005; @Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and species diversity of epiphytic and terrestrial fern communities.

@@

Moorea, French Polynesia (\ang{17;30;}S, \ang{149;50;}W) is a small (\SI{135}{\kilo\meter\squared}) tropical oceanic island more than \SI{5000}{\kilo\meter} from the nearest continental landmass.
This distance acts as a strong barrier to dispersal [@Carson1995; @Dassler2001], which combined with the young age of the island (\circa \SI{1.5}{\ma} [@Duncan1976]) has led to a relatively small, yet phylogenetically diverse fern flora (\circa \SI{130}{\species}, 8/11 orders; taxonomy follows @PteridophytePhylogenyGroupI2016; `r citest("species-list")`).
Ferns of Moorea include representatives of multiple Cretaceous epiphytic radiations (\eg, *Asplenium*, *Elaphoglossum*, Hymenophylloideae, Polypodiaceae, Vittarioideae; @Schuettpelz2009) and their terrestrial relatives, making them suitable for a comparative study of traits related to the evolution of epiphytic growth in ferns.

@@

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} $\times$ \SI{10}{\meter} plots spanning an elevational gradient from \SIrange{20}{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017].
Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}).
Each plot was divided into 25, \SI{2}{\meter} $\times$ \SI{2}{\meter} subplots.
Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots).
Epiphytic ferns were sampled by hand to \circa \SI{2}{\meter} on tree trunks or confirmed visually from the ground.
Species were categorized as epiphytic or terrestrial based on field observations (see Trait Selection and Measurements below).
The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as "epiphytic" and "terrestrial" communities.
Vouchers were deposited at GH and UC (abbreviations follow Index Herbariorum; @Thiers2019).

@@

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA).
A pair of dataloggers was installed for each plot, one mounted at \circa \SI{2}{\meter} on a tree ("epiphytic datalogger"), and one at ground level ("terrestrial datalogger"), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger.
Temperature and RH were logged once every \SI{15}{\minute} during two survey periods: a preliminary survey from `r prelim_climate_survey$start` to `r prelim_climate_survey$end` at `r prelim_climate_survey$n_sites %>% english2` plots, and a final survey from `r final_climate_survey$start` to `r final_climate_survey$end` at `r final_climate_survey$n_sites %>% english2` plots.
Two plots were not included in the environmental survey.
We used the 2013–-2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part (March 12--July 5, 2014) of the final survey period, so we used data for the same dates from the preliminary survey period for this datalogger instead.
Three others also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis.
Other days missing data during the final survey period were excluded from the dataset.
The final dataset included `r total_days` days of data for `r total_sites %>% filter(habit == "epiphytic") %>% pull(n)` epiphytic dataloggers and `r total_sites %>% filter(habit == "terrestrial") %>% pull(n)` terrestrial dataloggers.
We calculated vapor pressure deficit (VPD) from relative humidity and temperature using the \package{plantecophys} R package [@Duursma2015].

@@

We selected several traits relevant to epiphytic growth (`r citet("traits-used")` and `r citef("traits-used-image")`).
We preferred morphological traits over physiological traits because data for morphological traits are easier to gather at the scale of our study (\circa \SI{130}{\species} distributed over an elevational gradient from \SIrange{200}{2000}{\meter}).
Plant morphology and physiology are closely linked, and physiological traits such as rates of photosynthesis and water use efficiency are known to be correlated with morphological traits such as stipe length in ferns [@WatkinsJr.2010].

Morphological traits of sporophytes were measured from herbarium specimens, or obtained from the literature when plant material was unavailable.
Measurements were made directly on plants in the field for a few species with large fronds.
A single leaf or rhizome per individual was measured on one to three separately collected individuals per species, and mean values were calculated for each species.
Although trait values can vary within fern species with elevation [@Kessler2007a], our study includes a broad phylogenetic sampling, so we assume intraspecific variation is outweighed by interspecies differences.
To measure specific leaf area (SLA; \si{\square\meter\per\kilo\gram}), \circa 10--12, \SI{2}{\milli\meter} or \SI{4}{\milli\meter} diam. punches were taken of laminae between primary veins for a single leaf per individual using biopsy punches, dried at \SI{60}{\celsius} for \SI{24}{\hour}, and weighed.
For species with \textless \SI{2}{\milli\meter} between primary veins, leaf fragments excluding primary veins were obtained by dissection, dried at \SI{60}{\celsius} for \SI{24}{\hour}, weighed, and scanned.
ImageJ [@Abramoff2004] software was used to measure area per fragment.

Morphological traits of gametophytes were all qualitative (either categorical or binary; `r citet("traits-used")`).
Traits were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017] or grown from spores of known species in the lab, taken from the literature, or inferred from taxonomic group (i.e., genus) if documented as a stable taxonomic character.
For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate.
Non-cordate morphotypes include elongate (strap or ribbon \sensu @Farrar2008) and filamentous forms (`r citef("traits-used-image")`).
Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

@@

We calculated daily maximum, mean, minimum, and standard variation of temperature, RH, and VPD for each datalogger, then overall means from the daily values.
We analyzed correlation between the mean values with Pearson's correlation coefficient, and retained a set with correlation coefficients less than 0.9 (a single variable was arbitrarily selected from each set of correlated variables; hereafter, "climatic variables").
Variables relating to VPD, which have no meaning if less than zero, and some other variables that had skewed distributions, were square-root transformed prior to fitting models.
We tested for differences between epiphytic and terrestrial dataloggers while accounting for elevation with ANCOVA, using growth habit as the categorical variable and elevation as the covariate.
We identified the best-fitting linear model for each climatic variable in response to growth habit and elevation by constructing a set of models including the effect of each independent variable alone and their interactions, then selecting the one with the lowest corrected Akaike Information Criterion [AIC~c~; @Akaike1973].
A single outlier plot with unusually high temperature and low humidity (Mt. Rotui, 830 m plot) was excluded from ANCOVA and linear models.
Mt. Rotui is an isolated peak with greater exposure than the other two mountains where the rest of the plots were established (J. Nitta, pers. obs.).

@@

We tested for phylogenetic signal in quantitative traits using Blomberg's *K* [@Blomberg2003] and Pagel's λ [@Pagel1999], two metrics with slightly different interpretations and sensitivity to different evolutionary scenarios [@Munkemuller2012], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012].
λ is a scaling parameter that ranges from 0 (traits evolving randomly) to 1 (traits evolving along branches according to Brownian motion (BM)).
\K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under BM: when \K > 1, traits are more conserved than expected under BM; when \K < 1, traits have less phylogenetic signal than expected under BM [@Blomberg2003].
We tested for significance by comparing observed \K against a null distribution of \num{1000} trees with trait values randomized across the tips, and in λ by conducting a likelihood ratio test comparing the log likelihoods of the observed value of λ vs. λ = 0 (no phylogenetic signal).
We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' *D* [@Fritz2010].
Values of \D range from 0 under a model of trait evolution by BM, to 1 for random distribution of traits.
We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], and conducted significance test by comparing observed \D to distributions of simulated values produced under two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM model.
To account for possible circularity in measuring the phylogenetic signal of traits that were scored based on taxonomy, we conducted the same analyses on a dataset excluding species that had been scored this way ("strict" dataset).

@@

We detected significant phylogenetic signal in most traits, so we tested for differences between traits of epiphytic vs. terrestrial species using phylogenetic comparative methods.
For quantitative traits, we used the \function{brunch} function in the \package{caper} package [@Orme2013], which calculates phylogenetically independent contrasts in a quantitative character between alternative states of a binary trait.
For qualitative (binary) traits, we used Pagel's test of correlated evolution [@Pagel1994] as implemented with the \function{fitPagel} function in the \package{phytools} package [@Revell2012].
This method compares likelihoods between alternative models for a pair of binary traits.
In the first model, rates of evolution of the two traits are independent.
In the second model, the rate of evolution of one trait depends on the other.
A significantly better fit of the second model indicates that evolution of the two traits is correlated.
The same analyses were also done on the "strict" dataset.

We investigated differences in multivariate trait space between epiphytes and terrestrial species using principal components analysis (PCA).
We used only quantitative traits, and included only species that had no missing data (*n* = `r quant_traits_nsp` \si{\species}).
We first log-transformed traits as necessary to reduce skew, then scaled each trait by its standard deviation.
We conducted standard and phylogenetically corrected PCA, using the \function{PCA} and \function{phyl.pca} functions in the \package{FactoMineR} [@Le2008] and \package{phytools} [@Revell2012] R packages, respectively.

@@

Raw values of MPD and MNTD are statistically related to species richness [@Swenson2014]; to obtain metrics that are uncorrelated with species richness, we calculated their standard effect sizes (SES) by comparing observed values to a distribution of \num{999} randomized communities.
We constructed null communities using the "independent swap" algorithm [@Gotelli2000].
We included all species of ferns from Moorea and the neighboring island of Tahiti with genetic data available [`r Ntip(phy)` \si{\species} total; @Nitta2017] for phylogenetic null communities, and all species with trait data available (`r fern_traits %>%pull(species) %>% n_distinct`) for functional null communities.
We used the \function{ses.mpd} and \function{ses.mntd} functions in the R package \package{picante} [@Kembel2010] to calculate phylogenetic and functional diversity metrics.
Hereafter, "MPD~phy~" and "MNTD~phy~", and "MPD~func~" and "MNTD~func~", refer to the standard effect sizes of MPD and MNTD as measured with phylogenetic and functional trait distances, respectively.

We tested the effects of climate and growth habit on diversity metrics using full-subsets multiple regression with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018].
For each diversity metric, we built a set of models (Gaussian family, with plot as a random effect) taking into account all combinations of explanatory variables and their pairwise interactions, and ranked these by AIC~c~ to find the models that best explain the data.
Our dataset included `r (nrow(div_metrics_all) - nrow(climate_select)) %>% english2` communities that lacked climate data and could not be included in full-subsets multiple regression analysis.
Therefore, we also constructed linear models for each diversity metric related to elevation including all sites using the same model selection procedure as in the climate analysis.
We checked for spatial autocorrelation in the model residuals with Moran's *I* [@Moran1950] as implemented in the \package{spdep} R package [@Bivand2013].

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs `r table_nums("phylosig-cont", display = "n")`, `r table_nums("phylosig-binary", display = "n")`, `r citef("tree")`), and for quantitative traits, different results were obtained for λ and K.
When measured with λ, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of λ close to 1; only number of pinnae pairs had λ close to zero (`r citet("phylosig-cont")`).
However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of \K < 1 (`r citet("phylosig-cont")`).
All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D = `r hairs_d`) or more conserved (gemmae, \D = `r gemmae_d`; glands, \D = `r glands_d`; morphotype, \D = `r morphotype_d`) than expected under BM (`r citet("phylosig-binary")`).
Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D = `r habit_d`; `r citet("phylosig-binary")`).
Similar results were obtained using the "strict" dataset (`r citest("phylosig-binary-strict")`).

@@

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (`r citet("pics")`).
Epiphytic species tend to have smaller values for all traits (`r citet("pics")`; contrasts were made by subtracting epiphytic from terrestrial values).
Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, \pval `r stipe_pval`), as well as frond width (\pval `r width_pval`) and degree of laminae dissection (\pval `r dissection_pval`).
For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, \pval `r morphotype_pval`; `r citet("corr-evo")`).
Similar results were obtained using the "strict" dataset (`r citest("corr-evo-strict")`).

@@

**`r citef("traits-used-image")`**: Examples of traits used in this study. **(a)** Sporophyte traits. All sporophyte traits were quantitative, including frond length and width, stipe length, rhizome dia., number of pinnae pairs (in this example, nine) and degree of lamina dissection (in this example, once-pinnate). Specific leaf area (ratio of area to mass of leaf lamina) not depicted. **(b)** Gametophyte traits. Morphotype was defined as a binary trait, either cordate or non-cordate. Examples of non-cordate morphotypes include ribbon (thallus elongate and two-dimensional) and filamentous (thallus single lines of cells). Other gametophyte traits (gemmae, hairs, and glands) were scored as present or absent. Arrows point out instances of each binary trait. Scalebars 1 mm except for glands, which is 0.1 mm. Photographs by J. H. Nitta.

>>>>

<!-- Work-around to get pandoc to recognize custom siunitx units -->
\renewcommand{\hour}{h}
\renewcommand{\second}{s}
\renewcommand{\ma}{MA}
\renewcommand{\year}{yr}
\renewcommand{\month}{mo}
\renewcommand{\week}{wk}
\renewcommand{\day}{d}
\renewcommand{\basepair}{bp}

@@

```{r word-count table}
tibble(
  words = c(
    "Total word count (Introduction, Materials and Methods, Results, Discussion):_n",
    "Summary:_n",
    "Introduction:_n",
    "Materials and Methods:_n",
    "Results:_n",
    "Discussion:_n",
    "Acknowledgements:_n"),
  figs_tabs = c(
    "No. of figures:_7 (all in color)",
    "No. of tables:_5",
    "No. of supporting information files:_8 (Figs. S1-S2; Tables S1-S6)", "", "", "", "")
) %>%
  separate(words, c("words", "count"), sep = "_") %>%
  separate(figs_tabs, c("figs_tabs", "n"), sep = "_", fill = "right") %>%
  mutate(n = replace_na("")) %>%
  gt() %>%
  cols_label(
    words = "",
    count = "",
    figs_tabs = "",
    n = "",
  ) %>%
  as_latex
```

@@

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic fundamentally changed climates and had an immense effect on terrestrial ecosystems, including increased global precipitation and the creation of novel niches [@Niklas1983; @Kenrick1997a; @Lloyd2008; @Boyce2010]. 
The heterogeneous niche space of the angiosperm canopy comprising stratified light and humidity levels and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. 
The transition to epiphytic growth was particularly important for modern fern diversification: \circa{ }29% of all fern species are epiphytes vs. \circa{ }10% across all vascular plants [@Kress1986].
Moreover, major transitions to epiphytism occurred at least five times within ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012].
This has led to the hypothesis that epiphytism was a key factor contributing to extant fern diversification [@Schneider2004d; @Schuettpelz2009; @WatkinsJr.2012].
Recent studies exploring this idea [@Sundue2015; @Testo2016a; @Lehtonen2017], however, have largely failed to identify evidence of rapid diversification associated with epiphytism in ferns. 
This raises the possibility that modern fern diversification was not shaped by the evolution of epiphytism *per se*, but rather by the exploitation of various niches within the canopy and on the forest floor.

@@

To gain insight into the role of the canopy in promoting or restricting diversity, a functional and community-based approach can be applied to study this question. 
If the canopy is uniformly harsher than the terrestrial environment, thus acting as a 'filter', we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to increased functional diversity of epiphytic communities. 
Furthermore, the role of the canopy as a promoter or restrictor of diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kromer2005; @Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and species diversity of epiphytic and terrestrial fern communities.

@@

Moorea, French Polynesia (17°30'S, 149°50'W) is a small (\SI{135}{\kilo\meter}^2^) tropical oceanic island more than \SI{5000}{\kilo\meter} from the nearest continental landmass.
This distance acts as a strong barrier to dispersal [@Carson1995; @Dassler2001], which combined with the young age of the island (\circa{ }\SI{1.5}{\ma} [@Duncan1976]) has led to a relatively small, yet phylogenetically diverse fern flora (\circa{ }130 \species, 8/11 orders; taxonomy follows @PteridophytePhylogenyGroupI2016; `r citest("species-list")`).
Ferns of Moorea include representatives of multiple Cretaceous epiphytic radiations (\eg, *Asplenium*, *Elaphoglossum*, Hymenophylloideae, Polypodiaceae, Vittarioideae; @Schuettpelz2009) and their terrestrial relatives, making them suitable for a comparative study of traits related to the evolution of epiphytic growth in ferns.

@@

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} × \SI{10}{\meter} plots spanning an elevational gradient from \SI{20}{\meter} to \SI{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017].
Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}).
Each plot was divided into 25, \SI{2}{\meter} × \SI{2}{\meter} subplots.
Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots).
Epiphytic ferns were sampled by hand to \circa{ }\SI{2}{\meter} on tree trunks or confirmed visually from the ground.
Species were categorized as epiphytic or terrestrial based on field observations (see Trait Selection and Measurements below).
The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities.
Vouchers were deposited at GH and UC (abbreviations follow Index Herbariorum; @Thiers2019).

@@

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA).
A pair of dataloggers was installed for each plot, one mounted at \circa{ }\SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger.
Temperature and RH were logged once every \SI{15}{\minute} during two survey periods: a preliminary survey from `r prelim_climate_survey$start` to `r prelim_climate_survey$end` at `r prelim_climate_survey$n_sites %>% english2` plots, and a final survey from `r final_climate_survey$start` to `r final_climate_survey$end` at `r final_climate_survey$n_sites %>% english2` plots.
Two plots were not included in the environmental survey.
We used the 2013--2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part (March 12--July 5, 2014) of the final survey period, so we used data for the same dates from the preliminary survey period for this datalogger instead.
Three others also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis.
Other days missing data during the final survey period were excluded from the dataset.
The final dataset included `r total_days` days of data for `r total_sites %>% filter(habit == "epiphytic") %>% pull(n)` epiphytic dataloggers and `r total_sites %>% filter(habit == "terrestrial") %>% pull(n)` terrestrial dataloggers.
We calculated vapor pressure deficit (VPD) from relative humidity and temperature using the \package{plantecophys} R package [@Duursma2015].

@@

We selected several traits relevant to epiphytic growth (`r citet("traits-used")` and `r citef("traits-used-image")`).
We preferred morphological traits over physiological traits because data for morphological traits are easier to gather at the scale of our study (\circa{ }130 \species distributed over an elevational gradient from \SI{200}{\meter} to \SI{2000}{\meter}).
Plant morphology and physiology are closely linked, and physiological traits such as rates of photosynthesis and water use efficiency are known to be correlated with morphological traits such as stipe length in ferns [@WatkinsJr.2010].

Morphological traits of sporophytes were measured from herbarium specimens, or obtained from the literature when plant material was unavailable.
Measurements were made directly on plants in the field for a few species with large fronds.
A single leaf or rhizome per individual was measured on one to three separately collected individuals per species, and mean values were calculated for each species.
Although trait values can vary within fern species with elevation [@Kessler2007a], our study includes a broad phylogenetic sampling, so we assume intraspecific variation is outweighed by interspecies differences.
To measure specific leaf area (SLA; m^2^ kg^-1^), \circa{ }10--12, \SI{2}{\milli\meter} or \SI{4}{\milli\meter} \diameter punches were taken of laminae between primary veins for a single leaf per individual using biopsy punches, dried at \SI{60}{\celsius} for \SI{24}{\hour}, and weighed.
For species with < \SI{2}{\milli\meter} between primary veins, leaf fragments excluding primary veins were obtained by dissection, dried at \SI{60}{\celsius} for \SI{24}{\hour}, weighed, and scanned.
ImageJ [@Abramoff2004] software was used to measure area per fragment.

Morphological traits of gametophytes were all qualitative (either categorical or binary; `r citet("traits-used")`).
Traits were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017] or grown from spores of known species in the lab, taken from the literature, or inferred from taxonomic group (\ie, genus) if documented as a stable taxonomic character.
For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate [@Ebihara2013].
Non-cordate morphotypes include elongate (strap or ribbon \sensu{ }@Farrar2008) and filamentous forms (`r citef("traits-used-image")`).
Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

@@

We calculated daily maximum, mean, minimum, and standard variation of temperature, RH, and VPD for each datalogger, then overall means from the daily values.
We analyzed correlation between the mean values with Pearson's correlation coefficient, and retained a set with correlation coefficients less than 0.9 (a single variable was arbitrarily selected from each set of correlated variables; hereafter, 'climatic variables').
Variables relating to VPD, which have no meaning if less than zero, and some other variables that had skewed distributions, were square-root transformed prior to fitting models.
We tested for differences between epiphytic and terrestrial dataloggers while accounting for elevation with ANCOVA, using growth habit as the categorical variable and elevation as the covariate.
We identified the best-fitting linear model for each climatic variable in response to growth habit and elevation by constructing a set of models including the effect of each independent variable alone and their interactions, then selecting the one with the lowest corrected Akaike Information Criterion [AIC~c~; @Akaike1973].
A single outlier plot with unusually high temperature and low humidity (Mt. Rotui, 830 m plot) was excluded from ANCOVA and linear models.
Mt. Rotui is an isolated peak with greater exposure than the other two mountains where the rest of the plots were established (J. Nitta, pers. obs.).

@@

We tested for phylogenetic signal in quantitative traits using Blomberg's \K{ }[@Blomberg2003] and Pagel's \plam{ }[@Pagel1999], two metrics with slightly different interpretations and sensitivity to different evolutionary scenarios [@Munkemuller2012], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012].
\plam is a scaling parameter that ranges from 0 (traits evolving randomly) to 1 (traits evolving along branches according to Brownian motion (BM)).
\K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under BM: when \K{ }> 1, traits are more conserved than expected under BM; when \K{ }< 1, traits have less phylogenetic signal than expected under BM [@Blomberg2003].
We tested for significance by comparing observed \K against a null distribution of \num{1000} trees with trait values randomized across the tips, and in \plam by conducting a likelihood ratio test comparing the log likelihoods of the observed value of \plam vs. \plam{ }= 0 (no phylogenetic signal).
We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' \D{ }[@Fritz2010].
Values of \D range from 0 under a model of trait evolution by BM, to 1 for random distribution of traits.
We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], and conducted significance test by comparing observed \D to distributions of simulated values produced under two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM model.
To account for possible circularity in measuring the phylogenetic signal of traits that were scored based on taxonomy, we conducted the same analyses on a dataset excluding species that had been scored this way ('strict' dataset).

@@

We detected significant phylogenetic signal in most traits, so we tested for differences between traits of epiphytic vs. terrestrial species using phylogenetic comparative methods.
For quantitative traits, we used the \function{brunch} function in the \package{caper} package [@Orme2013], which calculates phylogenetically independent contrasts in a quantitative character between alternative states of a binary trait.
For qualitative (binary) traits, we used Pagel's test of correlated evolution [@Pagel1994] as implemented with the \function{fitPagel} function in the \package{phytools} package [@Revell2012].
This method compares likelihoods between alternative models for a pair of binary traits.
In the first model, rates of evolution of the two traits are independent.
In the second model, the rate of evolution of one trait depends on the other.
A significantly better fit of the second model indicates that evolution of the two traits is correlated.
The same analyses were also done on the 'strict' dataset.

We investigated differences in multivariate trait space between epiphytes and terrestrial species using principal components analysis (PCA).
We used only quantitative traits, and included only species that had no missing data (\n{ }= `r quant_traits_nsp` \species).
We first log-transformed traits as necessary to reduce skew, then scaled each trait by its standard deviation.
We conducted standard and phylogenetically corrected PCA, using the \function{PCA} and \function{phyl.pca} functions in the \package{FactoMineR} [@Le2008] and \package{phytools} [@Revell2012] R packages, respectively.

@@

Raw values of MPD and MNTD are statistically related to species richness [@Swenson2014]; to obtain metrics that are uncorrelated with species richness, we calculated their standard effect sizes (SES) by comparing observed values to a distribution of \num{999} randomized communities.
We constructed null communities using the 'independent swap' algorithm [@Gotelli2000].
We included all species of ferns from Moorea and the neighboring island of Tahiti with genetic data available [`r Ntip(phy)` \species total; @Nitta2017] for phylogenetic null communities, and all species with trait data available (`r fern_traits %>%pull(species) %>% n_distinct`) for functional null communities.
We used the \function{ses.mpd} and \function{ses.mntd} functions in the R package \package{picante} [@Kembel2010] to calculate phylogenetic and functional diversity metrics.
Hereafter, 'MPD~phy~' and 'MNTD~phy~', and 'MPD~func~' and 'MNTD~func~', refer to the standard effect sizes of MPD and MNTD as measured with phylogenetic and functional trait distances, respectively.

We tested the effects of climate and growth habit on diversity metrics using full-subsets multiple regression with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018].
For each diversity metric, we built a set of models (Gaussian family, with plot as a random effect) taking into account all combinations of explanatory variables and their pairwise interactions, and ranked these by AIC~c~ to find the models that best explain the data.
Our dataset included `r (nrow(div_metrics_all) - nrow(climate_select)) %>% jntools::english2` communities that lacked climate data and could not be included in full-subsets multiple regression analysis.
Therefore, we also constructed linear models for each diversity metric related to elevation including all sites using the same model selection procedure as in the climate analysis.
We checked for spatial autocorrelation in the model residuals with Moran's *I* [@Moran1950] as implemented in the \package{spdep} R package [@Bivand2013].

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs{ }`r table_nums("phylosig-cont", display = "n")`, `r table_nums("phylosig-binary", display = "n")`, `r citef("tree")`), and for quantitative traits, different results were obtained for \plam and \K.
When measured with \plam, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of \plam close to 1; only number of pinnae pairs had \plam close to zero (`r citet("phylosig-cont")`).
However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of \K{ }< 1 (`r citet("phylosig-cont")`).
All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D{ }= `r hairs_d`) or more conserved (gemmae, \D{ }= `r gemmae_d`; glands, \D{ }= `r glands_d`; morphotype, \D{ }= `r morphotype_d`) than expected under BM (`r citet("phylosig-binary")`).
Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D{ }= `r habit_d`; `r citet("phylosig-binary")`).
Similar results were obtained using the 'strict' dataset (`r citest("phylosig-binary-strict")`).

@@

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (`r citet("pics")`).
Epiphytic species tend to have smaller values for all traits (`r citet("pics")`; contrasts were made by subtracting epiphytic from terrestrial values).
Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, \pval{ }`r stipe_pval`), as well as frond width (\pval{ }`r width_pval`) and degree of laminae dissection (\pval{ }`r dissection_pval`).
For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, \pval{ }`r morphotype_pval`; `r citet("corr-evo")`).
Similar results were obtained using the 'strict' dataset (`r citest("corr-evo-strict")`).

@@

**`r citef("traits-used-image")`**: Examples of traits used in this study. **(a)** Sporophyte traits. All sporophyte traits were quantitative, including frond length and width, stipe length, rhizome dia., number of pinnae pairs (in this example, nine) and degree of lamina dissection (in this example, once-pinnate). Specific leaf area (ratio of area to mass of leaf lamina) not depicted. **(b)** Gametophyte traits. Morphotype was defined as a binary trait, either cordate or non-cordate. Examples of non-cordate morphotypes include ribbon (thallus elongate and two-dimensional) and filamentous (thallus single lines of cells). Other gametophyte traits (gemmae, hairs, and glands) were scored as present or absent. Arrows point out instances of each binary trait. Scalebars \SI{1}{\milli\meter} except for glands, which is \SI{.1}{\milli\meter}. Photographs by J. H. Nitta.
